Modeling HEPES buffer with a charge-balance approach

### 2023-11-18 manuscript.Rmd
Formatting tweaks

\begin{equation}
[\mathrm{H}^{+}] + SID - \frac{K_w}{[\mathrm{H}^{+}]} + \mathrm{I} \cdot 2
+ \mathrm{II} - \mathrm{IV} = 0
\end{equation}

@@

```{r}
pHm <- Vectorize(pH)
ans <- c(
 pH(.04, .05, -1, 3, 7.55),
 pH(.035, .05, -1, 3, 7.55)
)
all.equal(ans, pH.wrapper(
 c(.04, .035), .05, -1, 3, 7.55
))
all.equal(ans, pHm(
 c(.04, .035), .05, -1, 3, 7.55
))
```

@@

```{r}
pH.sym <- function(SID, HEPTOT, pK1, pK2, pK3) {
 H <- vapply(SID, function(SID) { # for every point...
  # find the root of the function
  H <- uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
                 HEPTOT = HEPTOT, SID = SID,
                 pK1 = pK1, pK2 = pK2, pK3 = pK3)$root
  # evaluate the derivatives by H and parameters
  g <- attr(eval(charge_balance), 'gradient')
  # and calculate the derivative of the H_eq by parameters;
  # combine together for easier access by vapply
  c(H, -g[,2:4]/g[,1])
 }, numeric(4)
 )
 # now take the negative logarithm to obtain pH
 structure(-log10(H[1,]), gradient = -(1/(H[1,] * log(10))) * t(H[-1,]))
}
pH.sym(
 c(.04, .035), .05, -1, 3, 7.55
)
```

This gets similar, but not exactly the same results as numeric
differentiation:

```{r}
nlsr::numericDerivR(
 quote(pHm(
  c(.04, .035), .05, pK1, pK2, pK3
 )),
 c('pK1', 'pK2', 'pK3'),
 list2env(list(pK1 = -1, pK2 = 3, pK3 = 7.55)),
 central = TRUE
)
```

>>>>

\begin{equation}
[\mathrm{H}^{+}] + \mathrm{SID} - \frac{K_w}{[\mathrm{H}^{+}]}
+ \mathrm{I} \cdot 2 + \mathrm{II} - \mathrm{IV} = 0
\end{equation}

@@

```{r}
ans <- c(
 pH(.04, .05, -1, 3, 7.55),
 pH(.035, .05, -1, 3, 7.55)
)
all.equal(ans, pH.wrapper(c(.04, .035), .05, -1, 3, 7.55))
pHm <- Vectorize(pH)
all.equal(ans, pHm(c(.04, .035), .05, -1, 3, 7.55))
```

@@

```{r}
pH.sym <- function(SID, HEPTOT, pK1, pK2, pK3) {
 H <- vapply(SID, function(SID) { # for every point:
  # find the root of the function
  H <- uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
                 HEPTOT = HEPTOT, SID = SID,
                 pK1 = pK1, pK2 = pK2, pK3 = pK3)$root
  # evaluate the derivatives by H and parameters
  g <- attr(eval(charge_balance), 'gradient')
  # and calculate the derivative of the H_eq by parameters;
  # combine together for easier access by vapply
  c(H, -g[,2:4]/g[,1])
 }, numeric(4))
 # now take the negative logarithm to obtain pH
 structure(-log10(H[1,]), gradient = -(1/(H[1,] * log(10))) * t(H[-1,]))
}
pH.sym(c(.04, .035), .05, -1, 3, 7.55)
```

This gives similar, but not exactly the same results as numeric
differentiation:

```{r}
nlsr::numericDerivR(
 quote(pHm(c(.04, .035), .05, pK1, pK2, pK3)),
 c('pK1', 'pK2', 'pK3'),
 list2env(list(pK1 = -1, pK2 = 3, pK3 = 7.55)),
 central = TRUE
)
```
